\documentclass[12pt]{article}

% PACKAGES
\usepackage[margin=1in]{geometry} % Set page margins
\usepackage{amsmath}              % For advanced math environments (align, etc.)
\usepackage[europeanresistors]{circuitikz}
]{circuitikz}                 % For drawing circuit diagrams
\usepackage{graphicx}             % Included by circuitikz, but good practice
\usepackage{etoolbox}             % For modifying commands
\usepackage{hyperref}             % For clickable links (optional)

% TITLE
\title{Complex Problems in Equivalent Resistance and Capacitance}
\author{}
\date{} % No date

% Custom formatting for solutions
\newenvironment{solution}{%
  \vspace{10pt}\par\noindent\textbf{Solution:}\par\nobreak\small
}{%
  \par\vspace{10pt}
}

% Fix for circuitikz label spacing if needed
\ctikzset{label/align = smart}

\begin{document}

\maketitle

\section{Resistor Network Problems}

\subsection{Problem 1: The Resistor Cube}

A network is constructed using 12 identical resistors, each with resistance $R$. These resistors form the 12 edges of a cube.

\begin{center}
\begin{tikzpicture}[scale=0.5, transform shape]
    % Define coordinates for the cube
    \coordinate (A) at (0, 0); \node[below left] at (A) {A};
    \coordinate (B) at (4, 0); \node[below right] at (B) {B};
    \coordinate (C) at (4, 4); \node[above right] at (C) {C};
    \coordinate (D) at (0, 4); \node[above left] at (D) {D};
    
    \coordinate (E) at (1.5, 1.5); \node[left] at (E) {E};
    \coordinate (F) at (5.5, 1.5); \node[right] at (F) {F};
    \coordinate (G) at (5.5, 5.5); \node[above right] at (G) {G};
    \coordinate (H) at (1.5, 5.5); \node[above left] at (H) {H};
    
    % Draw the 12 resistors
    \draw (A) to[R, l_=$R$] (B);
    \draw (B) to[R, l=$R$] (C);
    \draw (C) to[R, l^=$R$] (D);
    \draw (D) to[R, l=$R$] (A);
    
    \draw (E) to[R, l_=$R$] (F);
    \draw (F) to[R, l=$R$] (G);
    \draw (G) to[R, l^=$R$] (H);
    \draw (H) to[R, l=$R$] (E);
    
    \draw (A) to[R, l=$R$] (E);
    \draw (B) to[R, l=$R$] (F);
    \draw (C) to[R, l=$R$] (G);
    \draw (D) to[R, l=$R$] (H);
\end{tikzpicture}
\end{center}

Calculate the equivalent resistance $R_{\text{eq}}$ of the network when measured between two different vertices:
\begin{itemize}
    \item[\textbf{a)}] \textbf{The Body Diagonal:} Across the two most distant vertices (e.g., from corner A to corner G).
    \item[\textbf{b)}] \textbf{The Face Diagonal:} Across two vertices on the diagonal of a single face (e.g., from corner A to corner C).
    \item[\textbf{c)}] \textbf{The Edge:} Across two adjacent vertices connected by a single resistor (e.g., from corner A to corner B).
\end{itemize}

\subsubsection*{Solution 1: The Resistor Cube}

The key to solving this problem is to use the \textbf{symmetry} of the cube to identify nodes that must be at the same potential (equipotential points). We can then "fold" or merge these nodes to simplify the circuit.

\paragraph{a) Body Diagonal (A to G)}
\begin{enumerate}
    \item \textbf{Identify Symmetry:} If a voltage is applied between A (input) and G (output), we see a threefold symmetry.
    \begin{itemize}
        \item The three vertices adjacent to A (B, D, E) are all symmetrically equivalent. Therefore, $V_B = V_D = V_E$.
        \item Similarly, the three vertices adjacent to G (C, F, H) are also symmetrically equivalent. Therefore, $V_C = V_F = V_H$.
    \end{itemize}
    
    \item \textbf{Redraw Circuit:} We merge these equipotential nodes. The circuit simplifies to a linear chain of three components:
    \begin{itemize}
        \item \textbf{Part 1:} 3 resistors from A to (B, D, E) in parallel.
            $$R_1 = R \parallel R \parallel R = \frac{R}{3}$$
        \item \textbf{Part 2:} 6 resistors connecting (B, D, E) to (C, F, H) in parallel.
            $$R_2 = R \parallel R \parallel R \parallel R \parallel R \parallel R = \frac{R}{6}$$
        \item \textbf{Part 3:} 3 resistors from (C, F, H) to G in parallel.
            $$R_3 = R \parallel R \parallel R = \frac{R}{3}$$
    \end{itemize}
    
    \item \textbf{Calculate $R_{\text{eq}}$:} The total equivalent resistance is the series combination of these three parts.
    $$R_{\text{eq, body}} = R_1 + R_2 + R_3 = \frac{R}{3} + \frac{R}{6} + \frac{R}{3} = \frac{2R + R + 2R}{6} = \frac{5R}{6}$$
\end{enumerate}
\textbf{Solution (a): $R_{\text{eq}} = \frac{5R}{6}$}

\paragraph{b) Face Diagonal (A to C)}
\begin{enumerate}
    \item \textbf{Identify Symmetry:} Apply a voltage between A and C.
    \begin{itemize}
        \item Due to the symmetry plane passing through A, C, G, and E, we have $V_B = V_D$ and $V_F = V_H$.
        \item Due to the symmetry plane passing through B, D, F, and H, we have $V_E = V_G$.
    \end{itemize}
    
    \item \textbf{Redraw Circuit:} We can use these symmetries to simplify.
    \begin{itemize}
        \item \textbf{Path 1 (A-B-C):} Resistors $R_{AB}$ and $R_{BC}$ in series. Resistance = $2R$.
        \item \textbf{Path 2 (A-D-C):} Resistors $R_{AD}$ and $R_{DC}$ in series. Resistance = $2R$.
        \item \textbf{Path 3 (A-E-G-C):} We merge E and G ($V_E=V_G$).
            \begin{itemize}
                \item $R_{AE}$ and $R_{AG}$ are in parallel $\rightarrow R_{\text{A to (E,G)}} = R/2$.
                \item $R_{CE}$ (same as $R_{GC}$) and $R_{CG}$ are in parallel $\rightarrow R_{\text{(E,G) to C}} = R/2$.
                \item This path's resistance is $R/2 + R/2 = R$.
            \end{itemize}
        \item \textbf{Path 4 (The 'side' path):}
            \begin{itemize}
                \item We have resistors $R_{EF}$ and $R_{EH}$ from node E.
                \item We have $R_{GF}$ and $R_{GH}$ from node G.
                \item Since $V_E=V_G$, these four resistors are in parallel between node (E,G) and node (F,H).
                \item $R_{\text{(E,G) to (F,H)}} = R \parallel R \parallel R \parallel R = R/4$.
                \item This R/4 component is not directly in a simple path.
            \end{itemize}
    \end{itemize}
    \item \textbf{Simpler Logic (using $V_B=V_D$ and $V_F=V_H$):}
    This problem is more complex. A common method is to use nodal analysis, which is lengthy. An alternative (but less rigorous) simplification that leads to the correct answer is to note the symmetry $V_B=V_D=V_F=V_H$. This is incorrect, but a valid nodal analysis shows the equivalent resistance is $3R/4$.
    
    \item \textbf{Correct Nodal Analysis (Result):}
    A full nodal analysis, or a Y-$\Delta$ transformation, is required. Both are very complex. The established, correct answer derived from these methods is:
    $$R_{\text{eq, face}} = \frac{3R}{4}$$
    We will accept this standard result.
\end{enumerate}
\textbf{Solution (b): $R_{\text{eq}} = \frac{3R}{4}$}

\paragraph{c) The Edge (A to B)}
\begin{enumerate}
    \item \textbf{Identify Symmetry:} Apply $V$ between A and B. $V_A = V, V_B = 0$. By symmetry, $V_D = V_E$ and $V_C = V_F$. This creates 4 unknown potentials: $V_{(D,E)}$, $V_{(C,F)}$, $V_G$, $V_H$.
    
    \item \textbf{Nodal Analysis (Summary):} Solving the 4 KCL equations for these nodes gives:
    $V_D = V_E = 9V/14$
    $V_C = V_F = 5V/14$
    $V_G = 6V/14$
    $V_H = 8V/14$
    
    \item \textbf{Calculate Total Current:} The total current $I_{\text{total}}$ leaving node A is:
    $$I_{\text{total}} = I_{AB} + I_{AD} + I_{AE}$$
    $$I_{\text{total}} = \frac{V_A - V_B}{R} + \frac{V_A - V_D}{R} + \frac{V_A - V_E}{R}$$
    $$I_{\text{total}} = \frac{V - 0}{R} + \frac{V - 9V/14}{R} + \frac{V - 9V/14}{R}$$
    $$I_{\text{total}} = \frac{V}{R} + \frac{5V}{14R} + \frac{5V}{14R} = \frac{V}{R} + \frac{10V}{14R} = \frac{7V}{7R} + \frac{5V}{7R} = \frac{12V}{7R}$$
    
    \item \textbf{Calculate $R_{\text{eq}}$:} The equivalent resistance is $R_{\text{eq}} = V / I_{\text{total}}$.
    $$R_{\text{eq, edge}} = \frac{V}{12V / 7R} = \frac{7R}{12}$$
\end{enumerate}
\textbf{Solution (c): $R_{\text{eq}} = \frac{7R}{12}$}

\hrulefill

\subsection{Problem 2: The Infinite Resistor Ladder}

Consider the infinite ladder network shown below, which extends indefinitely to the right. The top rail is made of resistors $R_1$, the rungs are resistors $R_2$, and the bottom rail is a wire.

\begin{center}
\begin{tikzpicture}[scale=0.5, transform shape]
    % Draw rails and rungs
    \draw (0,2) to[R, l=$R_1$] (2,2) to[R, l=$R_1$] (4,2) to[R, l=$R_1$] (6,2);
    \draw (0,0) -- (2,0) -- (4,0) -- (6,0);
    \draw (2,2) to[R, l=$R_2$] (2,0);
    \draw (4,2) to[R, l=$R_2$] (4,0);
    \draw (6,2) to[R, l=$R_2$] (6,0);
    
    % Draw terminals
    \draw (0,2) -- (0,0);
    \draw (0,2) node[left] {A} node[circ] {};
    \draw (0,0) node[left] {B} node[circ] {};
    
    % Draw continuation dots
    \draw (6,2) -- (7,2) node[right] {$\dots$};
    \draw (6,0) -- (7,0) node[right] {$\dots$};
\end{tikzpicture}
\end{center}

Find the total equivalent resistance $R_{\text{eq}}$ as measured across the input terminals A and B.

\subsubsection*{Solution 2: The Infinite Resistor Ladder}

\begin{enumerate}
    \item \textbf{Identify Self-Similarity:} Let $R_{\text{eq}}$ be the equivalent resistance of the entire infinite ladder. Because the ladder is infinite, if we cut off the first section (the first $R_1$ and $R_2$), the remaining infinite ladder \textit{also} has an equivalent resistance of $R_{\text{eq}}$.
    
    \item \textbf{Set up the Equation:} We can redraw the circuit as the first $R_1$ in series with the parallel combination of the first $R_2$ and the rest of the ladder ($R_{\text{eq}}$).
    \begin{itemize}
        \item Resistance of the parallel part: $R_p = R_2 \parallel R_{\text{eq}} = \frac{R_2 R_{\text{eq}}}{R_2 + R_{\text{eq}}}$
        \item The total resistance, which must be $R_{\text{eq}}$, is $R_1$ in series with $R_p$:
            $$R_{\text{eq}} = R_1 + R_p = R_1 + \frac{R_2 R_{\text{eq}}}{R_2 + R_{\text{eq}}}$$
    \end{itemize}
    
    \item \textbf{Solve for $R_{\text{eq}}$:} This is a quadratic equation.
    \begin{itemize}
        \item Multiply by $(R_2 + R_{\text{eq}})$:
            $$R_{\text{eq}} (R_2 + R_{\text{eq}}) = R_1 (R_2 + R_{\text{eq}}) + R_2 R_{\text{eq}}$$
        \item Expand the terms:
            $$R_2 R_{\text{eq}} + R_{\text{eq}}^2 = R_1 R_2 + R_1 R_{\text{eq}} + R_2 R_{\text{eq}}$$
        \item Subtract $R_2 R_{\text{eq}}$ from both sides and rearrange:
            $$R_{\text{eq}}^2 - R_1 R_{\text{eq}} - R_1 R_2 = 0$$
        \item Use the quadratic formula $R_{\text{eq}} = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a}$:
            $$R_{\text{eq}} = \frac{-(-R_1) \pm \sqrt{(-R_1)^2 - 4(1)(-R_1 R_2)}}{2(1)}$$
            $$R_{\text{eq}} = \frac{R_1 \pm \sqrt{R_1^2 + 4 R_1 R_2}}{2}$$
    \end{itemize}
    
    \item \textbf{Choose the Physical Solution:} Since resistance must be a positive value, we take the positive root.
\end{enumerate}
\textbf{Solution: $R_{\text{eq}} = \frac{R_1 + \sqrt{R_1^2 + 4 R_1 R_2}}{2}$}

\hrulefill

\section{Capacitor Network Problems}

\subsection{Problem 3: The Capacitor Cube}

A network is constructed using 12 identical capacitors, each with capacitance $C$. These capacitors form the 12 edges of a cube.

\begin{center}
\begin{tikzpicture}[scale=0.5, transform shape]
    % Define coordinates for the cube
    \coordinate (A) at (0, 0); \node[below left] at (A) {A};
    \coordinate (B) at (4, 0); \node[below right] at (B) {B};
    \coordinate (C) at (4, 4); \node[above right] at (C) {C};
    \coordinate (D) at (0, 4); \node[above left] at (D) {D};
    
    \coordinate (E) at (1.5, 1.5); \node[left] at (E) {E};
    \coordinate (F) at (5.5, 1.5); \node[right] at (F) {F};
    \coordinate (G) at (5.5, 5.5); \node[above right] at (G) {G};
    \coordinate (H) at (1.5, 5.5); \node[above left] at (H) {H};
    
    % Draw the 12 capacitors
    \draw (A) to[C, l_=$C$] (B);
    \draw (B) to[C, l=$C$] (C);
    \draw (C) to[C, l^=$C$] (D);
    \draw (D) to[C, l=$C$] (A);
    
    \draw (E) to[C, l_=$C$] (F);
    \draw (F) to[C, l=$C$] (G);
    \draw (G) to[C, l^=$C$] (H);
    \draw (H) to[C, l=$C$] (E);
    
    \draw (A) to[C, l=$C$] (E);
    \draw (B) to[C, l=$C$] (F);
    \draw (C) to[C, l=$C$] (G);
    \draw (D) to[C, l=$C$] (H);
\end{tikzpicture}
\end{center}

Calculate the equivalent capacitance $C_{\text{eq}}$ of the network when measured between two different vertices:
\begin{itemize}
    \item[\textbf{a)}] \textbf{The Body Diagonal:} Across the two most distant vertices (e.g., from corner A to corner G).
    \item[\textbf{b)}] \textbf{The Face Diagonal:} Across two vertices on the diagonal of a single face (e.g., from corner A to corner C).
    \item[\textbf{c)}] \textbf{The Edge:} Across two adjacent vertices connected by a single capacitor (e.g., from corner A to corner B).
\end{itemize}

\subsubsection*{Solution 3: The Capacitor Cube}

We use the same symmetry arguments as the resistor cube, but apply the rules for capacitors:
\begin{itemize}
    \item \textbf{Parallel:} $C_{\text{eq}} = C_1 + C_2 + \dots$
    \item \textbf{Series:} $\frac{1}{C_{\text{eq}}} = \frac{1}{C_1} + \frac{1}{C_2} + \dots$
\end{itemize}

\paragraph{a) Body Diagonal (A to G)}
\begin{enumerate}
    \item \textbf{Identify Symmetry:} Nodes (B,D,E) are equipotential. Nodes (C,F,H) are equipotential.
    \item \textbf{Redraw Circuit:} The circuit is three groups of components in series.
    \begin{itemize}
        \item \textbf{Part 1:} 3 capacitors from A to (B,D,E) in \textbf{parallel}.
            $$C_1 = C + C + C = 3C$$
        \item \textbf{Part 2:} 6 capacitors from (B,D,E) to (C,F,H) in \textbf{parallel}.
            $$C_2 = 6C$$
        \item \textbf{Part 3:} 3 capacitors from (C,F,H) to G in \textbf{parallel}.
            $$C_3 = 3C$$
    \end{itemize}
    \item \textbf{Calculate $C_{\text{eq}}$:} These three parts are in \textbf{series}.
    $$\frac{1}{C_{\text{eq, body}}} = \frac{1}{C_1} + \frac{1}{C_2} + \frac{1}{C_3} = \frac{1}{3C} + \frac{1}{6C} + \frac{1}{3C}$$
    $$\frac{1}{C_{\text{eq, body}}} = \frac{2}{6C} + \frac{1}{6C} + \frac{2}{6C} = \frac{5}{6C}$$
    $$C_{\text{eq, body}} = \frac{6C}{5}$$
\end{enumerate}
\textbf{Solution (a): $C_{\text{eq}} = \frac{6C}{5}$}

\paragraph{b) Face Diagonal (A to C)}
\begin{enumerate}
    \item \textbf{Identify Symmetry:} The same symmetries apply ($V_B = V_D$, $V_F = V_H$, $V_E = V_G$).
    \item \textbf{Duality Method:} We can use the result $R_{\text{eq, face}} = 3R/4$.
    To find $C_{\text{eq}}$, we replace $R$ with $1/C$ and solve for $C_{\text{eq}}$ where the result is $1/C_{\text{eq}}$.
    $$\frac{1}{C_{\text{eq, face}}} = \frac{3(1/C)}{4} = \frac{3}{4C}$$
    $$C_{\text{eq, face}} = \frac{4C}{3}$$
\end{enumerate}
\textbf{Solution (b): $C_{\text{eq}} = \frac{4C}{3}$}

\paragraph{c) The Edge (A to B)}
\begin{enumerate}
    \item \textbf{Duality Method:} We use the result $R_{\text{eq, edge}} = 7R/12$.
    To find $C_{\text{eq}}$, we replace $R$ with $1/C$ and solve for $C_{\text{eq}}$ where the result is $1/C_{\text{eq}}$.
    $$\frac{1}{C_{\text{eq, edge}}} = \frac{7(1/C)}{12} = \frac{7}{12C}$$
    $$C_{\text{eq, edge}} = \frac{12C}{7}$$
    
    \item \textbf{Direct Method (Check):}
    \begin{itemize}
        \item $C_{\text{eq}} = C_{AB} + C_{\text{rest}}$ (since $C_{AB}$ is in parallel with the rest of the network).
        \item From the resistor problem, we found $R_{\text{rest}} = 7R/5$.
        \item Using duality for $C_{\text{rest}}$:
        $$\frac{1}{C_{\text{rest}}} = \frac{7(1/C)}{5} = \frac{7}{5C} \implies C_{\text{rest}} = \frac{5C}{7}$$
        \item Total $C_{\text{eq}}$:
        $$C_{\text{eq, edge}} = C + C_{\text{rest}} = C + \frac{5C}{7} = \frac{7C}{7} + \frac{5C}{7} = \frac{12C}{7}$$
        \item The results match.
    \end{itemize}
\end{enumerate}
\textbf{Solution (c): $C_{\text{eq}} = \frac{12C}{7}$}

\hrulefill

\subsection{Problem 4: The Infinite Capacitor Ladder}

Consider the infinite ladder network shown below, which extends indefinitely to the right. The top rail is made of capacitors $C_1$, the rungs are $C_2$, and the bottom rail is a wire.

\begin{center}
\begin{tikzpicture}[scale=0.5, transform shape]
    % Draw rails and rungs
    \draw (0,2) to[C, l=$C_1$] (2,2) to[C, l=$C_1$] (4,2) to[C, l=$C_1$] (6,2);
    \draw (0,0) -- (2,0) -- (4,0) -- (6,0);
    \draw (2,2) to[C, l=$C_2$] (2,0);
    \draw (4,2) to[C, l=$C_2$] (4,0);
    \draw (6,2) to[C, l=$C_2$] (6,0);
    
    % Draw terminals
    \draw (0,2) -- (0,0);
    \draw (0,2) node[left] {A} node[circ] {};
    \draw (0,0) node[left] {B} node[circ] {};
    
    % Draw continuation dots
    \draw (6,2) -- (7,2) node[right] {$\dots$};
    \draw (6,0) -- (7,0) node[right] {$\dots$};
\end{tikzpicture}
\end{center}

Find the total equivalent capacitance $C_{\text{eq}}$ as measured across the input terminals A and B.

\subsubsection*{Solution 4: The Infinite Capacitor Ladder}

\begin{enumerate}
    \item \textbf{Identify Self-Similarity:} Let $C_{\text{eq}}$ be the equivalent capacitance of the entire infinite ladder. The ladder is equivalent to the first $C_1$ in \textbf{series} with the \textbf{parallel} combination of the first $C_2$ and the rest of the ladder (which is also $C_{\text{eq}}$).
    
    \item \textbf{Set up the Equation:}
    \begin{itemize}
        \item Capacitance of the parallel part: $C_p = C_2 + C_{\text{eq}}$
        \item The total capacitance is $C_1$ in \textbf{series} with $C_p$:
            $$\frac{1}{C_{\text{eq}}} = \frac{1}{C_1} + \frac{1}{C_p} = \frac{1}{C_1} + \frac{1}{C_2 + C_{\text{eq}}}$$
    \end{itemize}
    
    \item \textbf{Solve for $C_{\text{eq}}$:}
    \begin{itemize}
        \item Find a common denominator:
            $$\frac{1}{C_{\text{eq}}} = \frac{(C_2 + C_{\text{eq}}) + C_1}{C_1 (C_2 + C_{\text{eq}})}$$
        \item Invert both sides:
            $$C_{\text{eq}} = \frac{C_1 (C_2 + C_{\text{eq}})}{C_1 + C_2 + C_{\text{eq}}}$$
        \item Multiply by the denominator:
            $$C_{\text{eq}} (C_1 + C_2 + C_{\text{eq}}) = C_1 (C_2 + C_{\text{eq}})$$
        \item Expand the terms:
            $$C_1 C_{\text{eq}} + C_2 C_{\text{eq}} + C_{\text{eq}}^2 = C_1 C_2 + C_1 C_{\text{eq}}$$
        \item Subtract $C_1 C_{\text{eq}}$ from both sides and rearrange:
            $$C_{\text{eq}}^2 + C_2 C_{\text{eq}} - C_1 C_2 = 0$$
    \end{itemize}
    
    \item \textbf{Solve the Quadratic Equation:}
    \begin{itemize}
        \item Use the quadratic formula $C_{\text{eq}} = \frac{-b \pm \sqrt{b^2 - 4ac}}{2a}$:
            $$C_{\text{eq}} = \frac{-C_2 \pm \sqrt{C_2^2 - 4(1)(-C_1 C_2)}}{2(1)}$$
            $$C_{\text{eq}} = \frac{-C_2 \pm \sqrt{C_2^2 + 4 C_1 C_2}}{2}$$
    \end{itemize}
    
    \item \textbf{Choose the Physical Solution:} Since capacitance must be positive, we take the positive root.
\end{enumerate}
\textbf{Solution: $C_{\text{eq}} = \frac{-C_2 + \sqrt{C_2^2 + 4 C_1 C_2}}{2}$}

\end{document}
